### 2021-05-14 docs/manuscript/PLOS-MSc-Thesis.tex
Remove \hil command for highlighting changes
Change tracking will be implemented with latexdiff.

% Highlight changes
\usepackage{color,soul}
\newcommand{\hil}[1]{\hl{#1}} % Highlight changes
% \newcommand{\hil}[1]{#1} % Do nothing, clean version

@@

Cells interact continuously with their environment and they must detect signals
from it. In many cases Gene Regulatory Networks (GRNs) interpret these signals
through gene interaction and produce a particular response. This information
processing function is of particular importance in developmental processes such
as those in which a group of cells differentiate in response to a signalling
molecule. Such molecules were referred to as morphogens by Turing in 1952
\cite{Turing1952} \hil{and posterior theoretical studies on patterning led to the
conceptualization of the French Flag Problem by Wolpert, who also stated
in this respect that a gradient of concentration of a morphogen could
trigger cell differentiation in a one-dimensional field of cells}
\cite{Wolpert1969, sharpe_2019}.
Although the Bicoid protein (bcd), the first example of a molecule that acted
as a morphogen, was only found in the 80's in the developing embryos of
\textit{Drosophila melanogaster}
\cite{driever_gradient_1988, driever_bicoid_1988}, now it is known that there
are many other examples such as the Decapentaplegic (Dpp) protein in
\textit{Drosophila} wing imaginal discs \cite{affolter_decapentaplegic_2007},
as well as Sonic Hedgehog \cite{dessaud_dynamic_2010,cohen_theoretical_2014}
and Wnt \cite{raspopovic_digit_2014}. However, information processing by GRNs
has proven to be a complex process, and a major goal of developmental biology
is to understand mechanistically how positional information given by morphogens
is translated in spatial differentiation.\\

@@

\hil{Cotterell and Sharpe studied what kind of 3-node network topologies could
interact with a morphogen gradient and produce a stripe of gene expression
in a one-dimensional field of cells and found a variety of networks that
implemented at least 7 different mechanisms and had different levels of
complexity} \cite{Cotterell2010}.

@@

The set of interaction values was represented as adjacency matrices $W \in
\mathbb{R}^{3\times4}$ where $i$ represents a regulated node of the GRN and $j$
represents a regulator node of the GRN (Fig \ref{fig:model}A and
\ref{fig:model}B) with $w_{\textit{ij}}~\in [-10,\ 10]$. Values in these
matrices can be negative, zero or positive and represent represion, no
interaction, or activation respectively. The magnitude of the value is
proportional to the strength of the interaction. \hil{In this study we define
a 'genotype' as the set of interaction, diffusion and degradation values that
represent a GRN.}\\

\begin{figure}[!h]
%  \centering
 \includegraphics[width=\textwidth]{figures/metodos/Fig1}
    \caption{\bf Modelling of GRNs.}
    (A) GRN represented as a directed graph with activation (blue arrows) and
    inhibition (red arrows) interactions. Black circles
    tagged with a \textbf{M} represent the morphogen; \textbf{A}, \textbf{B} and
    \textbf{C} represent genes in the network.
    (B) The same GRN represented as an adjacency matrix in which positive values
    represent activation, negative ones represent inhibition and
    zeros represent no interaction.
    (C) Morphogenetic field. In our model the field was composed of a linear
    array of 30 isogenic cells and a morphogen gradient concentration described
    by an exponential decay function. Horizontal arrows between cells represent
    diffusion of gene products between adjacent cells. Optimal gene expression
    pattern is shown top right.
    \hil{(D) Heatmap of spatiotemporal expression profile of a GRN where blue
    colors represent lower expression and red colors represent higher
    expression.}
 \label{fig:model}
\end{figure}

@@

\begin{figure}[!h]
 \includegraphics[width=\textwidth]{figures/results/Fig2}
 \caption{\bf Distribution of abundances of network topologies.}
 \hil{Some of the most} frequent network topologies shown in the grid
 contain the I3-FFL network motif \hil{(shown on the left). In this network
 motif the gene B is activated in an indirect way by the morphogen through
 the activation of A (positive interactions), and at the same time, it
 is inactivated in a direct way by the morphogen (negative interaction).}
 Topologies are numbered in order of decreasing abundance, \hil{being the
 Topology 1 the most abundant with a number of 95 genotypes}. The number of
 \hil{genotypes} belonging to each network topology is presented below
 topology number.
 Note that network topology number 9 produces the striped phenotype using a
 two-node GRN.
 \label{fig:distopol}
\end{figure}

@@

\begin{figure}[!h]
 \includegraphics[width=\textwidth]{figures/results/Fig3}
 \caption{\bf Neutral network.}
 \hil{This graph shows network topologies connected by one mutational step, where
 each node represents a single topology. Black nodes represent the most
 abundant topologies. All nodes with the same}
 color belong to a cluster of nodes more connected between them than with other
 nodes in the graph. \hil{The most abundant topologies are represented as
 directed graphs in the same way as in Fig}~\ref{fig:model}A, \hil{except for
 the topologies 1 to 8 and 15 (orange and red clusters), in which only
 topology 1 is depicted.}
 \label{fig:neutral-network}
\end{figure}

The degree distribution of this network seems to follow a Poisson distribution
(Pearson’s $\chi^2 = 12.8902$, $\text{p-value} = 0.115684$) characteristic of a
Erdös-Rényi network \cite{Erdos1959}. This kind of network presents a great
number of nodes with low degree and a few nodes with high degree
(Fig~\ref{fig:deg-dist}A). Nodes corresponding to topologies number one and
three showed the highest degree ($deg(1) = deg (3) = 14$). Curiously, nodes with
higher degree resulted to be also those network topologies with higher abundance
(Fig~\ref{fig:deg-dist}B). This relationship between abundance and node degree
was corroborated by the Spearman's rank correlation coefficient
(\hil{$r_S = 0.353752$}, $p = 6.2933\times10^{-23} $).

@@

\begin{figure}
 \includegraphics[width=\textwidth]{figures/results/Fig5}
 \caption{\bf Network of topologies separated by two mutational steps.}
 \hil{The network shows that almost all the topologies are connected by two or less
 mutational steps. Big red circle represents topology 420 that remains
 unconnected from the main graph. Big black circles correspond to the 15
 most abundant topologies described in} Fig~\ref{fig:neutral-network}.
 \hil{Different colors in the graph represent different clusters of topologies.}
 \label{fig:2neut-net}
\end{figure}

@@

\begin{figure}[!h]
 \includegraphics[width=\textwidth]{figures/results/Fig6}
 \caption{\bf Relationship between Abundance and robustness and complexity
 index.} (A) Abundance vs. Robustness. The robustness of each topology was
 calculated as the \hil{ratio} of perturbations in which the fitness was greater or
 equal to 0.95 with respect to the total of perturbations performed.
 (B) Abundance vs. Complexity index. Each point in the plot represents one of
 the 714 network topologies. \hil{More complex topologies tend to be less abundant
 than simple topologies
 ($r_S = -0.316653$, $\text{p-value} = 2.25161\times10^{-18}$).}
 \label{fig:ab-rob}
\end{figure}

@@

\begin{figure}[!h]
 \includegraphics[width=\textwidth]{figures/results/Fig8}
 \caption{\bf Expression dynamics of topology 9.}
 \hil{(A) Network topology number 9. Although this topology consists only of two
 genes, it display a striped pattern of gene expression for both of them as
 can be observed in (B). (B) Spatiotemporal expression profile of a GRN
 with topology 9. Gene \textbf{B} is not shown as it was lost in this topology
 in the evolutionary process simulated by the genetic algorithm.}
 \label{fig:topol9}
\end{figure}

@@

The correlation between the complexity index of a topology and its abundance
indicates that networks with simpler designs are more robust to changes in
interaction parameters and that more complex networks are less robusts. For
example, topology number 420 \hil{(Fig}~\ref{fig:2neut-net}), which
presents high complexity, was one of the
topologies with lower abundance and was disconnected from all other topologies
in the two-step neighbors in the neutral network.

>>>>

@@

Cells interact continuously with their environment and they must detect signals
from it. In many cases Gene Regulatory Networks (GRNs) interpret these signals
through gene interaction and produce a particular response. This information
processing function is of particular importance in developmental processes such
as those in which a group of cells differentiate in response to a signalling
molecule. Such molecules were referred to as morphogens by Turing in 1952
\cite{Turing1952} and posterior theoretical studies on patterning led to the
conceptualization of the French Flag Problem by Wolpert, who also stated
in this respect that a gradient of concentration of a morphogen could
trigger cell differentiation in a one-dimensional field of cells
\cite{Wolpert1969, sharpe_2019}.
Although the Bicoid protein (bcd), the first example of a molecule that acted
as a morphogen, was only found in the 80's in the developing embryos of
\textit{Drosophila melanogaster}
\cite{driever_gradient_1988, driever_bicoid_1988}, now it is known that there
are many other examples such as the Decapentaplegic (Dpp) protein in
\textit{Drosophila} wing imaginal discs \cite{affolter_decapentaplegic_2007},
as well as Sonic Hedgehog \cite{dessaud_dynamic_2010,cohen_theoretical_2014}
and Wnt \cite{raspopovic_digit_2014}. However, information processing by GRNs
has proven to be a complex process, and a major goal of developmental biology
is to understand mechanistically how positional information given by morphogens
is translated in spatial differentiation.\\

@@

Cotterell and Sharpe studied what kind of 3-node network topologies could
interact with a morphogen gradient and produce a stripe of gene expression
in a one-dimensional field of cells and found a variety of networks that
implemented at least 7 different mechanisms and had different levels of
complexity \cite{Cotterell2010}.

@@

The set of interaction values was represented as adjacency matrices $W \in
\mathbb{R}^{3\times4}$ where $i$ represents a regulated node of the GRN and $j$
represents a regulator node of the GRN (Fig \ref{fig:model}A and
\ref{fig:model}B) with $w_{\textit{ij}}~\in [-10,\ 10]$. Values in these
matrices can be negative, zero or positive and represent represion, no
interaction, or activation respectively. The magnitude of the value is
proportional to the strength of the interaction. In this study we define
a 'genotype' as the set of interaction, diffusion and degradation values that
represent a GRN.\\

\begin{figure}[!h]
%  \centering
 \includegraphics[width=\textwidth]{figures/metodos/Fig1}
    \caption{\bf Modelling of GRNs.}
    (A) GRN represented as a directed graph with activation (blue arrows) and
    inhibition (red arrows) interactions. Black circles
    tagged with a \textbf{M} represent the morphogen; \textbf{A}, \textbf{B} and
    \textbf{C} represent genes in the network.
    (B) The same GRN represented as an adjacency matrix in which positive values
    represent activation, negative ones represent inhibition and
    zeros represent no interaction.
    (C) Morphogenetic field. In our model the field was composed of a linear
    array of 30 isogenic cells and a morphogen gradient concentration described
    by an exponential decay function. Horizontal arrows between cells represent
    diffusion of gene products between adjacent cells. Optimal gene expression
    pattern is shown top right.
    (D) Heatmap of spatiotemporal expression profile of a GRN where blue
    colors represent lower expression and red colors represent higher
    expression.
 \label{fig:model}
\end{figure}

@@

\begin{figure}[!h]
 \includegraphics[width=\textwidth]{figures/results/Fig2}
 \caption{\bf Distribution of abundances of network topologies.}
 Some of the most frequent network topologies shown in the grid
 contain the I3-FFL network motif (shown on the left). In this network
 motif the gene B is activated in an indirect way by the morphogen through
 the activation of A (positive interactions), and at the same time, it
 is inactivated in a direct way by the morphogen (negative interaction).
 Topologies are numbered in order of decreasing abundance, being the
 Topology 1 the most abundant with a number of 95 genotypes. The number of
 genotypes belonging to each network topology is presented below
 topology number.
 Note that network topology number 9 produces the striped phenotype using a
 two-node GRN.
 \label{fig:distopol}
\end{figure}

@@

\begin{figure}[!h]
 \includegraphics[width=\textwidth]{figures/results/Fig3}
 \caption{\bf Neutral network.}
 This graph shows network topologies connected by one mutational step, where
 each node represents a single topology. Black nodes represent the most
 abundant topologies. All nodes with the same
 color belong to a cluster of nodes more connected between them than with other
 nodes in the graph. The most abundant topologies are represented as
 directed graphs in the same way as in Fig~\ref{fig:model}A, except for
 the topologies 1 to 8 and 15 (orange and red clusters), in which only
 topology 1 is depicted.
 \label{fig:neutral-network}
\end{figure}

The degree distribution of this network seems to follow a Poisson distribution
(Pearson’s $\chi^2 = 12.8902$, $\text{p-value} = 0.115684$) characteristic of a
Erdös-Rényi network \cite{Erdos1959}. This kind of network presents a great
number of nodes with low degree and a few nodes with high degree
(Fig~\ref{fig:deg-dist}A). Nodes corresponding to topologies number one and
three showed the highest degree ($deg(1) = deg (3) = 14$). Curiously, nodes with
higher degree resulted to be also those network topologies with higher abundance
(Fig~\ref{fig:deg-dist}B). This relationship between abundance and node degree
was corroborated by the Spearman's rank correlation coefficient
($r_S = 0.353752$, $p = 6.2933\times10^{-23} $).

@@

\begin{figure}
 \includegraphics[width=\textwidth]{figures/results/Fig5}
 \caption{\bf Network of topologies separated by two mutational steps.}
 The network shows that almost all the topologies are connected by two or less
 mutational steps. Big red circle represents topology 420 that remains
 unconnected from the main graph. Big black circles correspond to the 15
 most abundant topologies described in Fig~\ref{fig:neutral-network}.
 Different colors in the graph represent different clusters of topologies.
 \label{fig:2neut-net}
\end{figure}

@@

\begin{figure}[!h]
 \includegraphics[width=\textwidth]{figures/results/Fig6}
 \caption{\bf Relationship between Abundance and robustness and complexity
 index.} (A) Abundance vs. Robustness. The robustness of each topology was
 calculated as the ratio of perturbations in which the fitness was greater or
 equal to 0.95 with respect to the total of perturbations performed.
 (B) Abundance vs. Complexity index. Each point in the plot represents one of
 the 714 network topologies. More complex topologies tend to be less abundant
 than simple topologies
 ($r_S = -0.316653$, $\text{p-value} = 2.25161\times10^{-18}$).
 \label{fig:ab-rob}
\end{figure}

@@

\begin{figure}[!h]
 \includegraphics[width=\textwidth]{figures/results/Fig8}
 \caption{\bf Expression dynamics of topology 9.}
 (A) Network topology number 9. Although this topology consists only of two
 genes, it display a striped pattern of gene expression for both of them as
 can be observed in (B). (B) Spatiotemporal expression profile of a GRN
 with topology 9. Gene \textbf{B} is not shown as it was lost in this topology
 in the evolutionary process simulated by the genetic algorithm.
 \label{fig:topol9}
\end{figure}

@@

The correlation between the complexity index of a topology and its abundance
indicates that networks with simpler designs are more robust to changes in
interaction parameters and that more complex networks are less robusts. For
example, topology number 420 (Fig~\ref{fig:2neut-net}), which
presents high complexity, was one of the
topologies with lower abundance and was disconnected from all other topologies
in the two-step neighbors in the neutral network.
